Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-18 writing/paper-rbm/paper.tex
include new files

\documentclass[]{article}
%\usepackage{nips15submit_e,times}
\usepackage{url}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Properties and Bayesian fitting of restricted Boltzmann machines},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

% Tech stuff starts here
\usepackage{psfrag,epsf}
%\pdfminorversion=4

@@

%%INCLUDED-AK
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{remark}{Remark}

@@

\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
\bigskip

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\title{\bf Properties and Bayesian fitting of restricted Boltzmann machines}
\author{Andee Kaplan \\ Department of Statistical Science, Duke University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

\maketitle
\begin{abstract}
A restricted Boltzmann machine (RBM) is an undirected graphical model
constructed for discrete or continuous random variables, with two
layers, one hidden and one visible, and no conditional dependency within
a layer. In recent years, RBMs have risen to prominence due to their
connection to deep learning. By treating a hidden layer of one RBM as
the visible layer in a second RBM, a deep architecture can be created.
RBMs are thought to thereby have the ability to encode very complex and
rich structures in data, making them attractive for supervised learning.
However, the generative behavior of RBMs is largely unexplored and
typical fitting methodology does not easily allow for uncertainty
quantification in addition to point estimates. In this paper, we discuss
the relationship between RBM parameter specification in the binary case
and model properties such as degeneracy, instability and
uninterpretability. We also describe the associated difficulties that
can arise with likelihood-based inference and further discuss the
potential Bayes fitting of such (highly flexible) models, especially as
Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM
model structure.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

\hfill {\tiny technometrics tex template (do not remove)}
\newpage
\spacingset{1.45} % DON'T change the spacing!

@@

The data mining and machine learning communities have recently shown
great interest in deep learning, specifically in stacked restricted
Boltzmann machines (RBMs) (see Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009},
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008} for examples). A
RBM is a probabilistic undirected graphical model (for discrete or
continuous random variables) with two layers, one hidden and one
visible, with no conditional dependency within a layer (Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). These models
have reportedly been used with success in classification of images
(Larochelle and Bengio
\protect\hyperlink{ref-larochelle2008classification}{2008}; Srivastava
and Salakhutdinov
\protect\hyperlink{ref-srivastava2012multimodal}{2012}). However, the
model properties are largely unexplored in the literature and the
commonly cited fitting methodology remains heuristic-based and relies on
rough approximation (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). Additionally, the current
fitting methodology does not easily allow for interval estimation of
parameters in the model, and instead only provides for point estimation.
In this paper, we provide steps toward a fuller understanding of the
model class and its properties from the perspective of statistical model
theory, and we then explore the possibility of a rigorous fitting
methodology with natural uncertainty quantification. We find the RBM
model class to be concerning in two fundamental ways.

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect an observed data generation process. Such behavior relates to
model degeneracy (or near degeneracy), which is a statistical concern in
that many data processes of interest are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM used to model an imaging process, only a
small set of output possibilities receives nearly all probability, and
thus a sample of images will all be copies of the same one image (or
small number of images). An example of ten 4-pixel images simulated from
a nearly degenerate RBM model is compared to ten 4-pixel images
simulated from a non-degenerate RBM model in Figure
\ref{fig:sample-models}. The degenerate model places almost all
probability on one outcome, causing the image to be generated
repeatedly, whereas the non-degenerate model shows more potentially
realistic variation.

\par
\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}
In addition to such degeneracy, we find that RBMs can easily exhibit
types of instability, related to how sensitive the model probabilities
can be to small difference in data outcomes. In practice, this may be
seen when a single pixel change in an image results in a wildly
different classification in an image classification problem. Occurrences
of such behavior have recently been documented in RBMs (Li
\protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
architectures (Szegedy et al.
\protect\hyperlink{ref-szegedy2013intriguing}{2013}; Nguyen, Yosinski,
and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
(related) potential issues of model instability, degeneracy and
uninterpretability for the RBM class (which are related properties) and
examine the presence of these in Section
\ref{explorations-of-model-properties-through-simulation} through
simulations of small, manageable examples.

Separate from the exploration of model properties, we also investigate
the quality of estimation possible in fitting these models and examine
an estimation procedure that allows for uncertainty quantification
rather than point estimates alone. The fitting of RBMs can be
problematic for two reasons, the first being computational and the
second concerning flexibility. As the size of these models grows, both
maximum likelihood and Bayesian methods of fitting quickly become
intractable. The literature often suggests Markov chain Monte Carlo
(MCMC) tools for approximate maximization of likelihoods to fit these
models (e.g., Gibbs sampling to exploit conditional structure in hidden
and visible variables), but little is said about the attributes of
realized estimates (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Hinton, Osindero, and
Teh \protect\hyperlink{ref-hinton2006fast}{2006}), including the absence
of interval estimates.

@@

As a factor compounding the computational challenges, issues in model
fitting traceable to model flexibility are potentially more concerning.
For a RBM model with enough hidden variables, any distribution for the
visibles can be approximated arbitrarily well (Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008}; Montufar and Ay
\protect\hyperlink{ref-montufar2011refinements}{2011}; and Montúfar,
Rauh, and Ay \protect\hyperlink{ref-montufar2011expressive}{2011}).
However, for binary data, the empirical distribution of an observed
training set of visible variables provides a highest likelihood
benchmark, before any parametric model class is even introduced and
applied to obtain a refinement of model fit. As a consequence, we find
that any fully principled fitting method based on the likelihood for a
RBM with enough hidden variables will seek to reproduce the (discrete)
empirical distribution of a training set. This aspect can be
undesirable, and perhaps even unexpected compared to most modeling
scenarios, in that no ``smoothed distribution'' may result when fitting
a RBM model of sufficient size with a rigorous likelihood-based method.
We are therefore led to be skeptical that models that involve these
structures (like deep Boltzmann machines) can achieve useful prediction
or inference in a principled way without intentionally limiting the
flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
The degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the standard overall model-specification-and-fitting methodology
(Hinton \protect\hyperlink{ref-hinton2002training}{2002};
Carreira-Perpinan and Hinton
\protect\hyperlink{ref-carreira2005contrastive}{2005}). In this paper,
we attempt to address the concerns with overfitting and poor mixing in
an alternative (and perhaps more transparent) manner, via specification
of a Bayesian prior in Section \ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. This work aligns with the call to action by Google
scientists in Sculley et al.
(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
examples to advance empirical understanding of deep learning models and
strive for a ``higher level of empirical rigor in the field.'' We
examine three Bayesian fitting techniques intended to avoid model
impropriety issues raised in Section \ref{model-fitting} and conclude
with a discussion in Section \ref{discussion}. Supplementary online
material provides proofs for results on RBM parameterizations and data
codings described in Section \ref{data-encoding}.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, binary variables
being most commonly considered. In this paper, we consider the binary
case for concreteness. A RBM architecture has two layers, hidden
(\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with no dependency
connections within a layer. An example of this structure is in Figure
\ref{fig:rbm} with the hidden nodes indicated by gray circles and the
visible nodes indicated by white circles.
\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}
A common use for RBMs is to create features for use in classification.
For example, binary images can be classified through a process that
treats the pixel values as the visible variables \(v_i\) in a RBM model
(Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}).

@@

Due to the potential size of the model, the normalizing constant
\(\gamma(\boldsymbol \theta)\) can be practically impossible to
calculate, making simple estimation of the model parameter vector
problematic. In model fitting, a kind of Gibbs sampling can be tried due
to the conditional architecture of the RBM (i.e.~visibles given hiddens
or vice verse). Specifically, the conditional independence of nodes in
each layer (given those nodes in the other layer) allows for stepwise
simulation of both hidden layers and model parameters (e.g., see the
contrastive divergence of Hinton
(\protect\hyperlink{ref-hinton2002training}{2002}) or Bayes methods in
Section \ref{model-fitting}).

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the methodology claim to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution with a probability mass function
of the form \begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*} where
\(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the model
parameter, and \(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\)
is a vector of statistics based on the adjacency matrix of a graph.
Here, as earlier for RBMs,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parameterized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

>>>>

\documentclass[AMS,STIX1COL]{WileyNJD-v2}

\articletype{Research Article}

\received{17 September 2018}

\raggedbottom

\usepackage{amsmath}
% \usepackage{amsthm}
% \usepackage{cite}
\usepackage{hyperref}
\usepackage{graphics}
\usepackage{subfig}
\usepackage{url}

@@

%%INCLUDED-AK
% \usepackage{amsthm}
% \newtheorem{theorem}{Theorem}
% \newtheorem{proposition}{Proposition}
% \newtheorem{lemma}{Lemma}
% \newtheorem{corollary}{Corollary}
% 
% \theoremstyle{definition}
% \newtheorem{definition}{Definition}
% \newtheorem{assumption}{Assumption}
% \newtheorem{remark}{Remark}

@@

\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }

%% rmarkdown being obstinant?
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%hacks
%https://tex.stackexchange.com/questions/371955/hyperref-error-argument-of-hysetreflink-has-an-extra/384969
\AtBeginDocument{
  \label{FirstPage2}
  \def\fpage{\pageref{FirstPage2}}
}

\begin{document}

\title{Properties and Bayesian fitting of restricted Boltzmann machines}
\authormark{KAPLAN \textsc{et al}}

\author[1]{Andee Kaplan*}
\author[2]{Daniel Nordman}
\author[2,3]{Stephen Vardeman}

\address[1]{
\orgdiv{Department of Statistical Science}, \orgname{Duke University}, \orgaddress{\state{North Carolina}, \country{USA}}}
\address[2]{
\orgdiv{Department of Statistics}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{USA}}}
\address[3]{
\orgdiv{Department of Industrial and Manufacturing Systems Engineering}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{USA}}}

\corres{*Andee Kaplan, Department of Statistical Science, Duke University, P.O. Box 90251,
Durham, NC 27708-0251. \email{\href{mailto:andrea.kaplan@duke.edu}{\nolinkurl{andrea.kaplan@duke.edu}}}}

\abstract[Summary]{
A restricted Boltzmann machine (RBM) is an undirected graphical model
constructed for discrete or continuous random variables, with two
layers, one hidden and one visible, and no conditional dependency within
a layer. In recent years, RBMs have risen to prominence due to their
connection to deep learning. By treating a hidden layer of one RBM as
the visible layer in a second RBM, a deep architecture can be created.
RBMs are thought to thereby have the ability to encode very complex and
rich structures in data, making them attractive for supervised learning.
However, the generative behavior of RBMs is largely unexplored and
typical fitting methodology does not easily allow for uncertainty
quantification in addition to point estimates. In this paper, we discuss
the relationship between RBM parameter specification in the binary case
and model properties such as degeneracy, instability and
uninterpretability. We also describe the associated difficulties that
can arise with likelihood-based inference and further discuss the
potential Bayes fitting of such (highly flexible) models, especially as
Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM
model structure.
}
\keywords{Degeneracy, Instability, Classification, Deep Learning, Graphical Models}

\jnlcitation{\cname{
\author{Kaplan A.,} 
\author{D. Nordman, and} 
\author{S. Vardeman} 
 (\cyear{2018+}), 
\ctitle{Properties and Bayesian fitting of restricted Boltzmann machines}, \cjournal{Stat Anal DataMin: The ASA Data Sci Journal}, \cvol{2018+;00:00-00}.}}

\maketitle

@@

The data mining and machine learning communities have recently shown
great interest in deep learning, specifically in stacked restricted
Boltzmann machines (RBMs) (see Salakhutdinov and Hinton 2009, 2012;
Srivastava, Salakhutdinov, and Hinton 2013; Le Roux and Bengio 2008 for
examples). A RBM is a probabilistic undirected graphical model (for
discrete or continuous random variables) with two layers, one hidden and
one visible, with no conditional dependency within a layer (Smolensky
1986). These models have reportedly been used with success in
classification of images (Larochelle and Bengio 2008; Srivastava and
Salakhutdinov 2012). However, the model properties are largely
unexplored in the literature and the commonly cited fitting methodology
remains heuristic-based and relies on rough approximation (Hinton,
Osindero, and Teh 2006). Additionally, the current fitting methodology
does not easily allow for interval estimation of parameters in the
model, and instead only provides for point estimation. In this paper, we
provide steps toward a fuller understanding of the model class and its
properties from the perspective of statistical model theory, and we then
explore the possibility of a rigorous fitting methodology with natural
uncertainty quantification. We find the RBM model class to be concerning
in two fundamental ways.

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher (1922), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann 1990; G. E. P.
Box 1967). At issue, simulation from RBMs can often produce data lacking
realistic variability so that such models may thereby fail to
satisfactorily reflect an observed data generation process. Such
behavior relates to model degeneracy (or near degeneracy), which is a
statistical concern in that many data processes of interest are
realistically not degenerate in their spectrum of potential outcomes.
For example, when sampling data from a nearly degenerate RBM used to
model an imaging process, only a small set of output possibilities
receives nearly all probability, and thus a sample of images will all be
copies of the same one image (or small number of images). An example of
ten 4-pixel images simulated from a nearly degenerate RBM model is
compared to ten 4-pixel images simulated from a non-degenerate RBM model
in Figure \ref{fig:sample-models}. The degenerate model places almost
all probability on one outcome, causing the image to be generated
repeatedly, whereas the non-degenerate model shows more potentially
realistic variation.

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}

In addition to such degeneracy, we find that RBMs can easily exhibit
types of instability, related to how sensitive the model probabilities
can be to small difference in data outcomes. In practice, this may be
seen when a single pixel change in an image results in a wildly
different classification in an image classification problem. Occurrences
of such behavior have recently been documented in RBMs (Li 2014), as
well as other deep architectures (Szegedy et al. 2013; Nguyen, Yosinski,
and Clune 2014). We describe (related) potential issues of model
instability, degeneracy and uninterpretability for the RBM class (which
are related properties) and examine the presence of these in Section
\ref{explorations-of-model-properties-through-simulation} through
simulations of small, manageable examples.

Separate from the exploration of model properties, we also investigate
the quality of estimation possible in fitting these models and examine
an estimation procedure that allows for uncertainty quantification
rather than point estimates alone. The fitting of RBMs can be
problematic for two reasons, the first being computational and the
second concerning flexibility. As the size of these models grows, both
maximum likelihood and Bayesian methods of fitting quickly become
intractable. The literature often suggests Markov chain Monte Carlo
(MCMC) tools for approximate maximization of likelihoods to fit these
models (e.g., Gibbs sampling to exploit conditional structure in hidden
and visible variables), but little is said about the attributes of
realized estimates (Hinton 2010; Hinton, Osindero, and Teh 2006),
including the absence of interval estimates.

@@

As a factor compounding the computational challenges, issues in model
fitting traceable to model flexibility are potentially more concerning.
For a RBM model with enough hidden variables, any distribution for the
visibles can be approximated arbitrarily well (Le Roux and Bengio 2008;
Montufar and Ay 2011; and Montúfar, Rauh, and Ay 2011). However, for
binary data, the empirical distribution of an observed training set of
visible variables provides a highest likelihood benchmark, before any
parametric model class is even introduced and applied to obtain a
refinement of model fit. As a consequence, we find that any fully
principled fitting method based on the likelihood for a RBM with enough
hidden variables will seek to reproduce the (discrete) empirical
distribution of a training set. This aspect can be undesirable, and
perhaps even unexpected compared to most modeling scenarios, in that no
``smoothed distribution'' may result when fitting a RBM model of
sufficient size with a rigorous likelihood-based method. We are
therefore led to be skeptical that models that involve these structures
(like deep Boltzmann machines) can achieve useful prediction or
inference in a principled way without intentionally limiting the
flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
2010; Tieleman 2008; Cho, Ilin, and Raiko 2012). Both \(L_1\) and
\(L_2\) type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
The degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the standard overall model-specification-and-fitting methodology
(Hinton 2002; Carreira-Perpinan and Hinton 2005). In this paper, we
attempt to address the concerns with overfitting and poor mixing in an
alternative (and perhaps more transparent) manner, via specification of
a Bayesian prior in Section \ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. This work aligns with the call to action by Google
scientists in Sculley et al. (2018) to employ small toy examples to
advance empirical understanding of deep learning models and strive for a
``higher level of empirical rigor in the field.'' We examine three
Bayesian fitting techniques intended to avoid model impropriety issues
raised in Section \ref{model-fitting} and conclude with a discussion in
Section \ref{discussion}. Supplementary online material provides proofs
for results on RBM parameterizations and data codings described in
Section \ref{data-encoding}.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, binary variables
being most commonly considered. In this paper, we consider the binary
case for concreteness. A RBM architecture has two layers, hidden
(\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with no dependency
connections within a layer. An example of this structure is in Figure
\ref{fig:rbm} with the hidden nodes indicated by gray circles and the
visible nodes indicated by white circles.

\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}

A common use for RBMs is to create features for use in classification.
For example, binary images can be classified through a process that
treats the pixel values as the visible variables \(v_i\) in a RBM model
(Hinton, Osindero, and Teh 2006).

@@

Due to the potential size of the model, the normalizing constant
\(\gamma(\boldsymbol \theta)\) can be practically impossible to
calculate, making simple estimation of the model parameter vector
problematic. In model fitting, a kind of Gibbs sampling can be tried due
to the conditional architecture of the RBM (i.e.~visibles given hiddens
or vice verse). Specifically, the conditional independence of nodes in
each layer (given those nodes in the other layer) allows for stepwise
simulation of both hidden layers and model parameters (e.g., see the
contrastive divergence of Hinton (2002) or Bayes methods in Section
\ref{model-fitting}).

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh 2006; Salakhutdinov and
Hinton 2012; Srivastava, Salakhutdinov, and Hinton 2013 for examples).
By stacking multiple layers of RBMs in a deep architecture, proponents
of the methodology claim to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(Salakhutdinov and Hinton 2009, 450). The stacking is achieved by
treating a hidden layer of one RBM as the visible layer in a second RBM,
and so on, until the desired multi-layer architecture is created.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock 2003). For
random graph models, \(\mathcal{X}\) denotes all possible graphs that
can be constructed from a set of nodes and an exponentially
parameterized random graph model has a distribution with a probability
mass function of the form \begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*} where
\(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the model
parameter, and \(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\)
is a vector of statistics based on the adjacency matrix of a graph.
Here, as earlier for RBMs,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock (2003) classifies an exponentially
parameterized random graph model at \(\boldsymbol \theta\) as
\emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (2003), when the random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from \eqref{eq:t}, appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), and the RBM
model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is \begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}

\hypertarget{instability}{%
\subsubsection{Instability}\label{instability}}
